---
title: 'Stewart artificial selection lineages: thermal Plasticity experiment'
author: "Ian Dworkin"
date: "`r format(Sys.time(),'%d %b %Y')`"
output:
  html_document:
    keep_md: yes
    number_sections: yes
    toc: yes
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 3)
```


## Load libraries
```{r}
require(car)
require(emmeans)
require(ggplot2)
require(lme4)
require(glmmTMB)
require(lmerTest)
require(effects)
require(predictmeans)
require(ggbeeswarm)
require(ggdist)
require(ggridges)
require(cowplot)
```


```{r}
pref_theme <- theme_classic() +
  theme(text = element_text(size = 14))
```


## load data and do quality checks

```{r}
size_dat <- read.csv("../data/AS_LotithTempPlasticity.csv", h = T,
                     stringsAsFactors = TRUE)

str(size_dat)
levels(size_dat$Treat)

head(size_dat)

with(size_dat, table(Sex, Treat, Temp))
```


### check the class of some objects so they are what we need for analysis (i.e. factors instead of integers)

sex, selection, replicate, sampling, trait, individual_ID and repeat_measure should all be factors.

```{r}
str(size_dat)
names(size_dat)

size_dat$TempF <- size_dat$Temp # Temperature, coded as a factor.
names(size_dat)

# use lapply to convert strings to factors where we need them
size_dat[, c(3,4,7)] <- lapply(size_dat[, c(3,4,7)], factor)
str(size_dat) # check that it behaves as expected

size_dat$Treat <- relevel(size_dat$Treat, "LHm")
```



## log transform size

I may not use it, but we should always check.

From $mm$ to $\mu m$ and then $log_2$ transformed
```{r}
size_dat$log2_thorax <- log(1000*size_dat$Thorax, base = 2)
```


labels
```{r}
levels(size_dat$Treat) <- c("LHm", "Control", "Reversal", "Large", "Small")
  
treat.labs <- c(expression(LH^M), "Control", "Reversal", "Large", "Small")

names(treat.labs) <- c(levels(size_dat$Treat))
```



### Now check some basic data characteristics

Let's make sure we don't have any really obvious outliers (like something 10 times too big or ten times too small). Also make sure mean and median are pretty close.

```{r}
with(size_dat,
     tapply(Thorax, interaction(Treat, TempF, Sex, drop = T ), quantile, 
            probs = c(0,0.005, 0.02, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95,0.98,0.99, 1)))

with(size_dat,
     tapply(Thorax, interaction(Treat, TempF, Sex, drop = T ), mean))
```



We could also just look at min and max for each as a check.
```{r}
with(size_dat, tapply(Thorax,
                      interaction(Treat, TempF, Sex, drop = T ), 
                      range))
```




### graphical summaries of the raw data

I am providing a few different ways of plotting this. You can choose which you think is best to make your points.

These all show the same thing, just in different ways. Mostly a matter of preference depending on what you want to highlight in your paper.

```{r}
ggplot(size_dat, aes( y = Thorax, x = TempF, col = Treat:Sex)) +
  geom_boxplot(outlier.shape = NA) + 
  geom_point(size = 1.25, alpha = 0.25, 
                position=position_jitterdodge(0.5), show.legend= FALSE) +
  labs( y = "length (mm)") +
  theme_classic() +
  theme(text = element_text(size = 24))
```


```{r}
# Same as above faceted by sex
ggplot(size_dat, aes( y = Thorax, x = Treat, col = TempF)) +
  geom_boxplot(outlier.shape = NA) + 
  geom_point(size = 1.25, alpha = 0.25, 
                position=position_jitterdodge(0.5), show.legend= FALSE) +
  labs( y = "length (mm)") +
  facet_wrap(~ Sex) +
  theme_classic() +
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 45, size = 12, vjust = 0.6))
```


```{r}
# log transformed
ggplot(size_dat, aes( y = log2_thorax, x = Treat, col = TempF)) +
  geom_boxplot(outlier.shape = NA) + 
  geom_point(size = 1.25, alpha = 0.25, 
                position=position_jitterdodge(0.5), show.legend= FALSE) +
  labs( y = "thorax length log2 (µm)") +
  facet_wrap(~ Sex) +
  theme_classic() +
  theme(text = element_text(size = 24),
        axis.text.x = element_text(angle = 45, size = 12, vjust = 0.6))
```


```{r}
# violin plots as an alternative. Using log2 transformed. 
ggplot(size_dat, aes( y = log2_thorax, col = Treat, x = TempF)) +
  geom_violin(aes(fill = Treat)) +
  labs( y = "length, log2 (µm)") +
  facet_wrap(~ Sex) +
  theme_classic() +
  theme(text = element_text(size = 20))
```

Or beeswarm

```{r}
ggplot(size_dat, aes( y = log2_thorax, col = Treat, x = TempF)) +
  geom_quasirandom(alpha = 0.5) +
  labs(y = "length, log2 (µm)") +
  facet_wrap(~ Sex) +
  theme_classic() +
  theme(text = element_text(size = 20))
```


```{r}
ggplot(size_dat, aes( y = log2_thorax, x = Treat, col = TempF, shape = TempF)) +
  geom_quasirandom(alpha = 0.5) +
  labs(y = "length, log2 (µm)") +
  facet_wrap(~ Sex) +
  theme_classic() +
  theme(text = element_text(size = 20),
        axis.text.x = element_text(angle = 45, size = 12, vjust = 0.6))
```


```{r}
ggplot(size_dat, aes( y = log2_thorax, x = TempF, col = Sex, shape = Sex)) +
  geom_quasirandom(alpha = 0.5, method = "tukeyDense") +
  labs(#y = "length, log2 (µm)", 
       y = expression(thorax~length~log[2]~mu~m),
       x = "Temperature (°C)") +
  theme_bw() +
  facet_wrap(~ Treat, labeller = labeller(Treat = treat.labs)) +
  theme(text = element_text(size = 15),
        legend.position = "none")
```


## Additional plots of raw data to examine variation by vials and replicate block


```{r}
ggplot(size_dat, 
       aes( y = Thorax, x = Treat, col = TempF:Sex)) +
  geom_boxplot(outlier.shape = NA) + 
  geom_point(size = 1.25, alpha = 0.25, 
                position=position_jitterdodge(0.5), show.legend= FALSE) +
  labs( y = "length (mm)") +
  facet_wrap(~ Rep) +
  ggtitle("thorax") +
  theme_classic() +
  theme(text = element_text(size = 20),,
        axis.text.x = element_text(angle = 45, size = 12, vjust = 0.6))
```



### Real models (these have the info you want)

 temperature as a continuous predictor, where 0 represents the lowest tempereature used in the experiment.
 
```{r}
mean(size_dat$Temp)
size_dat$TempC <- size_dat$Temp - min(size_dat$Temp)
mean(size_dat$TempC)

size_dat$TempC_2 <- size_dat$Temp - mean(size_dat$Temp)
```


## About the analysis


Replicate Block is a bit hard to decide on whether to fit as fixed or random. It is obviously a nuisance variable, but with only two blocks it is not easy to estimate. I did include it as a random effect

Vial is nested within the joint levels of Treatment, TempF and Rep



## for the check of the evolved differences 
from the normal (what they evolved at) rearing temperature of 25°C


```{r}
mod1_thorax__25C_log <- lmer(log2(1000*Thorax) ~ Sex*Treat +
                    + (1 | Rep) 
                    + (1| Rep:Treat:Vial),
           REML = FALSE,          
           data = size_dat,
           subset = size_dat$Temp == 25)

summary(mod1_thorax__25C_log)
```


```{r}
estimates_25C <- emmeans(mod1_thorax__25C_log, ~ Treat | Sex)

estimates_25C

contrasts_To_control <- contrast(estimates_25C, method = "trt.vs.ctrl", ref = "Control")

contrasts_To_control

confint(contrasts_To_control)
```

please note, to recapitulate figures from the paper, remove the `type = "response"` from the emmeans statement so it will be displayed on the log2 scale. 

first panel (p1)
```{r}
p1 <- plot(contrasts_To_control) + 
  geom_vline(xintercept = 0, lty = 2, alpha = 0.5) +
  xlab(expression(paste("∆thorax length, ", log[2], " µm"))) + 
  ylab("") +
  theme_bw() +
  theme(text = element_text(size = 15), 
        legend.position = "none",
        legend.title = element_blank())
  

p1
```


using log2 µm, (which will allow us to focus on proportional changes.)

```{r}
mod1_thorax_log <- lmer(log2(1000*Thorax) ~ Sex*Treat*TempC +
                    + (1 | Rep) 
                    + (1 | Rep:Treat:TempF:Vial),
                    REML = FALSE,  
                    data = size_dat )

summary(mod1_thorax_log)

Anova(mod1_thorax_log)
```



Generate model fitted values and their CIs
```{r}
pred_vals_slopes <- as.data.frame(predictorEffect("TempC", mod1_thorax_log, 
                                                  x.levels = 13))

pred_vals_slopes$Temp <- pred_vals_slopes$TempC + 18

pred_vals_slopes$log2_thorax <- pred_vals_slopes$fit
```


### panel b

```{r}
pred_vals_slopes$Treat <- relevel(pred_vals_slopes$Treat, "LHm")
```


```{r}
p2 <- ggplot(data = pred_vals_slopes, aes(x = Temp, y = log2_thorax, colour = Sex)) +
  facet_wrap(~Treat, nrow = 1) +
  geom_line() +
  geom_ribbon(aes(ymin = pred_vals_slopes$lower, 
                  ymax = pred_vals_slopes$upper,
                  x = Temp,
                  colour = Sex), alpha = 0.175, 
               linetype = 0) +
  geom_jitter(data = size_dat,
              aes(y = log2_thorax, x = Temp), alpha = 0.2, width = 1) +
  ylab(expression(paste("Thorax length, ", log[2], " µm"))) + 
  xlab("Rearing temperature (°C)") + 
  theme_bw() +
  theme(text = element_text(size = 15), 
        legend.position = "top",
        legend.title = element_blank())


p2
```



## Changes in slopes


### What is the magnitude of change in the slope of the plasticity response, by experimental treatment.

Note that in the paper, the figures show the contrast computed on the log2 scale. 

```{r}
slopes2 <- emtrends(mod1_thorax_log,  var = "TempC",  
                            specs = ~ Treat | Sex)


plot(slopes2) + 
  labs(x = "Plasticity", y = "") +
  theme_bw() +
  theme(text = element_text(size = 18))

# relative to LHm
slope_contrasts_LHm <- contrast(slopes2, 
                             "trt.vs.ctrl", ref = "LHm",
                                 by = "Sex")

#excluding LHm, relative to controls
slope_contrasts_control <- contrast(slopes2, 
                             "trt.vs.ctrl", ref = "Control", exclude = 1, 
                                 by = "Sex")

confint(slope_contrasts_control)
```


```{r}
p3 <- plot(slope_contrasts_control) + 
  geom_vline(xintercept = 0, lty = 2, alpha = 0.5) + 
  labs(x = "∆plasticity relative to control", y = "") +
  scale_x_continuous(n.breaks = 4) + 
  theme_bw() +
  theme(text = element_text(size = 15), 
        legend.position = "none",
        legend.title = element_blank())

p3
```


```{r}
plot(slope_contrasts_LHm) + 
  geom_vline(xintercept = 0, lty = 2, alpha = 0.5) + 
  labs(x = "∆plasticity relative to LHm", y = "") +
  theme_bw() +
  theme(text = element_text(size = 15))
```

Just to help understand what the output is telling us
```{r}
mod1_thorax <- lmer(1000*Thorax ~ Sex*Treat*TempC +
                    + (1 | Rep) 
                    + (1| Rep:Treat:TempF:Vial),
                    REML = FALSE,
                    data = size_dat )


slopes2_b <- emtrends(mod1_thorax,  var = "TempC",  
                            specs = ~ Treat | Sex)

slopes2_b 


slope_contrasts_control_b <- contrast(slopes2_b,
                             "trt.vs.ctrl", ref = "Control", exclude = 1,
                                 by = "Sex")

confint(slope_contrasts_control_b)
```


## stitching together for the plot

```{r}
top_row <- plot_grid(p1, p3, 
          label_size = 20,        
          labels = c("A", "C"))

fig_paper <- plot_grid(top_row, p2,
          labels = c("", "B"),
          label_size = 20,
          ncol = 1)

fig_paper


save_plot("../outputs/fig_paper_file.png", fig_paper, base_height = 6.72, base_asp = 5/3, dpi = 1000)

save_plot("../outputs/fig_paper_file.tiff", fig_paper, base_height = 6.72, base_asp = 5/3, dpi = 500 )
```

## Note below, this is for ID's personal interest in changes in SSD, not about the paper!

The analysis below is because I was curious... Not about the paper *per se*
### compute differences in SSD for slopes slope relative to control

```{r}
slope_treatment <- emtrends(mod1_thorax_log,  var = "TempC",  
                            pairwise ~ Sex*Treat)

SSD_for_slope_contrasts <- contrast(slope_treatment[[1]], 
                                 interaction = c(Treat = "trt.vs.ctrl1", Sex = "pairwise"),
                                 by = NULL)



confint(SSD_for_slope_contrasts)

plot(SSD_for_slope_contrasts) + 
  geom_vline(xintercept = 0, lty = 2, alpha = 0.5) + 
  labs(x = "change in dimorphism plasticity relative to control lineages", y = "comparison") +
  theme_bw()
```


```{r}
sessionInfo()
```

